---
title: "[unit name], [chapter name] - Code snippets"
output: html_document
---

# [unit name], [chapter name] - Code examples

The following code relates to the **[chapter name]** section of the **[unit name]** learning resource on the [analysis for action](link/to/be/added) platform.  

Use this document to explore and run the code snippets provided in the learning resource.

## Troubleshooting

If you encounter any issues running the code snippets, consider the following troubleshooting steps:
- Ensure you have followed the setup instructions in the [README.md](../../../README.md) file including `install.packages()` for required packages.
- Run code from top to bottom to ensure all dependencies are met.
- Restart your R session and clear all outputs.
- For more detailed troubleshooting guidance, refer to the [troubleshooting document](LINK/TO/BE/ADDED).

## Setup

Before running the code snippets, ensure you have:
- followed the setup instructions in the [README.md](../../../README.md) file,
- **[add further instructions as needed]**,
- run the code block below.

```{r setup, message=FALSE}
# This chunk should include:
# * library() calls for required packages
# * paths to data files
# * any other setup required
```

## [Section 1 name]

[add a brief description of what the code is doing]

```{r section1}
# Add your code for Section 1 here
```

## [Section 2 name]

[add a brief description of what the code is doing]

```{r section2}
# Add your code for Section 2 here
```

[add more sections as needed]

---
