---
title: "[unit name], [chapter name] - Code snippets"
output: html_document
---

---
This block contains instructions on how to fill in this template.

**Please delete this block when creating your own file.**

This template should be used to create code snippet notebooks for each chapter
in a learning resource. Follow the instructions below to complete the template:
- Replace content in square brackets (e.g., [unit_name], [link], [chapter name])
with the relevant information. For example the title block below could be changed
from "[unit_name], [chapter_name] - Code snippets" to "Basic Data Visualisation,
Creating Bar Charts - Code snippets".
- In the setup section add any instructions for users to follow before running the
code snippets, such as installing packages or downloading data files. Add any
necessary import statements or setup code in the first code cell.
- Code snippets should be grouped into sections according to the sections in the
learning resource.
- Each code snippet from the learning resource should be placed in its own code
cell under the relevant section, with any necessary context provided in a
markdown block above each code cell.
- Follow [tidyverse](https://style.tidyverse.org/) style guide for R code
formatting.
- Whenever you need to reference a file in [learning_resources](http://_vscodecontentref_/1), use:
```r
file_path <- file.path(learning_resources_dir, "data", "filename.csv")
```

- Definitions:
    - "Unit name" refers to the unit name, which is also the name of the folder
    this python folder is in.
    - "Chapter name" refers to the specific chapter that the code snippets are
    in. Chapters are the highest level headings in the learning resource.
    - "Section name" refers to subheadings within chapters that code snippets
    are in.
---

# [unit name], [chapter name] - Code examples

The following code relates to the **[chapter name]** section of the
**[unit name]** learning resource on the [Analysis for Action]([PLACEHOLDER])
platform.

Use this document to explore and run the code snippets provided in the learning
resource.

## Troubleshooting

If you encounter any issues running the code snippets, consider the following
troubleshooting steps:
- Ensure you have followed the setup instructions in the
[README.md](../../../README.md) file including `install.packages()` for
required packages.
- Run code from top to bottom to ensure all dependencies are met.
- Restart your R session and clear all outputs.
- For more detailed troubleshooting guidance, refer to the
[troubleshooting document]([PLACEHOLDER]).

## Setup

Before running the code snippets, ensure you have:
- followed the setup instructions in the [README.md](../../../README.md) file,
- **[add further instructions as needed]**,
- run the code block below.

```{r setup, message=FALSE}
# This chunk should include:
# * library() calls for required packages
# * paths to data files
# * any other setup required
library(here)

# Use here::here() to construct paths, do not setwd()
learning_resources_dir <- here::here("learning_resources")
print(paste("Learning resources directory:", learning_resources_dir))
```

## [Section 1 name]

[add a brief description of what the code below is doing]

```{r section1a}
# Add the first code snippet from the first section of the relevant learning resource here
```

[add a brief description of what the code below is doing]
```{r section1b}
# Add the next code snippet from the first section of the relevant learning resource here
```

## [Section 2 name]

[add a brief description of what the code below is doing]

```{r section2a}
# Add your code for Section 2 here
```

[add more sections as needed]

---
