---
title: "Basic Data Visualisation, How to make bar charts in R"
output: html_document
---

This block contains instructions on how to fill in this template.

**Please delete this block when creating your own file.**

This template should be used to create code snippet notebooks for each chapter in a learning resource. Follow the instructions below to complete the template: - Replace content in square brackets (e.g., [unit_name], [link], [chapter name]) with the relevant information. For example the title block below could be changed from "[unit_name], [chapter_name] - Code snippets" to "Basic Data Visualisation, Creating Bar Charts - Code snippets". - In the setup section add any instructions for users to follow before running the code snippets, such as installing packages or downloading data files. Add any necessary import statements or setup code in the first code cell. - Code snippets should be grouped into sections according to the sections in the learning resource. - Each code snippet from the learning resource should be placed in its own code cell under the relevant section, with any necessary context provided in a markdown block above each code cell. - Follow [tidyverse](https://style.tidyverse.org/) style guide for R code formatting. - Whenever you need to reference a file in [learning_resources](http://_vscodecontentref_/1), use:

``` r
file_path <- file.path(learning_resources_dir, "data", "filename.csv")
```

-   Definitions:
    -   "Unit name" refers to the unit name, which is also the name of the folder this python folder is in.
    -   "Chapter name" refers to the specific chapter that the code snippets are in. Chapters are the highest level headings in the learning resource.
    -   "Section name" refers to subheadings within chapters that code snippets are in. ---

## Basic Data Visualisation, How to make bar charts in R - Code examples

The following code relates to the **How to make bar charts in R** section of the **Basic Data Visualisation** learning resource on the [Analysis for Action](%5BPLACEHOLDER%5D) platform.

Use this code to get to grips with the basics of making tables in R as per the best practice covered in the written guidance. You are welcome to copy and adapt the code to meet your needs, but make sure that any outputs also align with any style requirements for your organisation.

The code applies each change separately and does not create any outputs.

## Troubleshooting

If you encounter any issues running the code snippets, consider the following troubleshooting steps: - Ensure you have followed the setup instructions in the [README.md](../../../README.md) file including `install.packages()` for required packages. - Run code from top to bottom to ensure all dependencies are met. - Restart your R session and clear all outputs. - For more detailed troubleshooting guidance, refer to the [troubleshooting document](%5BPLACEHOLDER%5D).

## Setup

Before running the code snippets, ensure you have: - followed the setup instructions in the [README.md](../../../README.md) file, - **[add further instructions as needed]**, - run the code block below.

```{r setup, message=FALSE}
# This chunk should include:
# * library() calls for required packages
# * paths to data files
# * any other setup required

library(readr)
library(here)
library(dplyr)
library(tidyr)
library(janitor)
library(ggplot2)
library(svglite)
library(scales)

# Use here::here() to construct paths, do not setwd()
here::here()
vuln <- read.csv(here::here("learning_resources", "data", "vulnerable.csv"))
```

## Introduction to ggplot2

The main package you will use to create a table in this file is ggplot2.
The package works using layers, creating the base chart and then adding layers to change titles, colours and more.

## Prepare your data

This code uses the 'vulnerable' dataset.

The raw data is too large to use in a bar chart. You need to filter it to make a smaller dataframe first.

```{r}
# Prepare your data to make it suitable for a bar chart

vuln_2007_top_6 <- vuln %>%
  # Filter the data to just the year 2007
  filter(year %in% ("2007")) %>%
  # Select just the country and vulnerable_pop columns
  select(country, vulnerable_pop) %>%
  # Select only the rows with the 6 highest vulnerable_pop value
  dplyr::top_n(n = 6)

# To Display the data

vuln_2007_top_6
```

## Make the base bar chart

```{r}
# Creating a gglot2 bar chart

vuln_bar <- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity")

vuln_bar
```

## Make the chart horizontal

```{r}
vuln_bar_horizontal <- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity") +
  # Flip the chart so that the bars are horizontal
  coord_flip()

vuln_bar_horizontal
```

## Rename axes

```{r}
vuln_bar_titles <- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity") +
  # Add axis titles
  labs(x = "Country", y = "Vulnerable Population")
  
vuln_bar_titles
```

## Set a maximum value for the y-axis

```{r}
vuln_bar_y_axis_changed <- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity") +
  # Set a limit for the y-axis
  ylim(0, 400000000)

vuln_bar_y_axis_changed
```

## Reorder columns

```{r}
vuln_2007_top_6_arranged <- vuln_2007_top_6 %>% 
  # Order the data frame rows from largest to smallest
  dplyr::arrange(desc(vulnerable_pop))

# dplyr::arrange(vulnerable_pop) to reverse the order

vuln_2007_top_6_arranged$country <- factor(
  vuln_2007_top_6_arranged$country,
  levels = vuln_2007_top_6_arranged$country
)

# Create the bar chart using the arrngede data
vuln_bar_arranged <- ggplot(vuln_2007_top_6_arranged, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity") 

vuln_bar_arranged
```

## Change bar width

```{r}
vuln_bar_change_width <- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity",
           # Change the bar width
           width = 0.6)

vuln_bar_change_width
```

## Change colours of bars

```{r}
vuln_bar_change_bar_colour <- vuln_bar <- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity", fill = "#324027")

vuln_bar_change_bar_colour
```

## Change bar colours (focus chart)

```{r}
# Create a new DataFrame which is a copy of vuln_2007_top_6 with a column called highlight which contains the value TRUE if the row contains the maximum vulnerable_pop value, all other rows contain the value FALSE in the highlight column
vuln_focus <- vuln_2007_top_6 %>%
  mutate(highlight = vulnerable_pop == max(vulnerable_pop))

# Create a bar chart with the vuln_focus data 
vuln_bar_focus_chart <- ggplot(vuln_focus, aes(x = country, y = vulnerable_pop, fill = highlight)) +
  geom_bar(stat = "identity") +
  # Change the bar colours based off the whether highlight is TRUE or FALSE
  scale_fill_manual(values = c("TRUE" = "#1e2a16", "FALSE" = "#959595"),
                    guide = "none")
vuln_bar_focus_chart
```

## Add labels to bars

```{r}
vuln_bar_add_labels <- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity") +
  # Add labels to bars containing the vulnerable population of each country
  geom_text(aes(label = vulnerable_pop), 
            # Adjust the position of the label
            vjust = -0.25,
            # Adjust the text size of the label
            size = 4)  

vuln_bar_add_labels
```

## Specify the number of axis break

```{r}
vuln_bar_custom_axis_breaks <- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity") +
  # Custom axis breaks are used, y-axis is breaks are every 50000000 people
  scale_y_continuous(breaks = c(0, 50000000, 100000000, 150000000, 200000000, 250000000, 300000000, 350000000),
  # The function expansion is used to increase the vertical height of the plot by 40% to show the extra axis breaks
  expand = expansion(mult = c(0, 0.4)))
                                
vuln_bar_custom_axis_breaks 
```

## Change background and gridlines

```{r}
vuln_bar_change_background_and_gridlines<- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity", 
           # Change the colour of the bars to ensure contrast with grey gridlines
           fill = "#48553f") +
  theme(
    # Change the background colour of the panel (the inside box containing the bars) to white
    panel.background = element_rect(fill = "#ffffff"),
    # Change the background colour of the plot (the outside area containing the axis titles)
    plot.background = element_rect(fill = "#f4f6f5"),
    # Changed size of axis titles and axis text.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(size = 8),
    axis.text.y = element_text(size = 8),
    # Change the colour and size of the minor gridlines
    panel.grid.minor = element_line(colour = "#444443", size = 0.5, linetype = 1),
    # Change the colour and size of the major gridlines
    panel.grid.major = element_line(colour = "#444443", size = 0.5, linetype = 1))
    # Note that line types after linetype = 1 are different versions of dotted and dashed gridlines
  

vuln_bar_change_background_and_gridlines
```

## Remove gridlines

```{r}
vuln_bar_no_gridlines <- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity") +
  # Remove gridlines
  theme(panel.grid = element_blank())

vuln_bar_no_gridlines
```

# Add number suffixes (Thousands)

```{r}
vuln_bar_thousands <- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = label_number(scale = 1e-3, suffix = "k")) +
  labs(y = "Vulnerable Population (Thousands)")

vuln_bar_thousands
```

# Add number suffixes (Millions)

```{r}
vuln_bar_millions <- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = label_number(scale = 1e-6, suffix = "M")) +
  labs(y = "Vulnerable Population (Millions)")

vuln_bar_millions
```

# Add number suffixes (Billions)

```{r}
vuln_bar_billions <- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = label_number(scale = 1e-9, suffix = "B")) +
  labs(y = "Vulnerable Population (Billions)")

vuln_bar_billions
```

## Show figures in full

```{r}
vuln_bar_figures_in_full <- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = scales::label_number(accuracy = 1))
vuln_bar_figures_in_full
```

## Exporting

```{r}
vuln_bar <- ggplot(vuln_2007_top_6, aes(x = country ,y = vulnerable_pop)) +
  geom_bar(stat = "identity")

ggsave(
  # Set the file name
  filename = "vuln_pop_2007.svg",
  # Select the plot to save
  plot = vuln_bar,
  # Set the save path
  path = "D:/Temp",
  # Set aspect ratio of plot, an aspect ratio of 16:9 will fill 1920x1080 displays
  width = 16,
  height = 9
)
```

## Exercise

Use this exercise to test what you’ve learned about making bar charts. You can adapt much of the code covered in this section, but some basic data manipulation will also be required which has not been covered here.

Create a bar chart in R. Use the vulnerable dataset as your input data.

For an extra challenge, consider setting yourself an hour’s time limit to help simulate an emergency.

The bar chart should:

-   Show the size of vulnerable populations in Argentina, Malawi, Nepal, and the United Kingdom in the year 1997
-   Be formatted so that:
    -   The x-axis label is: Country 
    -   The y-axis label is: Vulnerable Population (Millions) 
    -   The text size of the axis titles is 12
    -   The panel back ground is HEX code #ffffff
    -   The plot background is HEX code #f3f6f4
    -   The largest bar is HEX code #324028, the rest are HEX code #D4DDD6
    -   The width of the bars is 0.6
    -   The bars are ordered from largest to smallest
    -   The horizontal and vertical gridlines are both HEX code #444443 and size = 0.2
    -   The y-axis uses millions suffix (for example an axis break with a value of
        6000000 would show as 6M)
    -   The chart is exported as an .svg file with the file name:
        vuln_pop_partner_countries 
    -   The chart is saved to the D:/Temp folder
    -   The save charted has an aspect ratio of 16:9

When you have finished making your bar chart, compare it against the relevant code and solution in the 'exercise_solutions' subfolder. If you have followed all the guidance correctly, this is approximately how your output should look.

## Extras

## Clustered Bar Chart

```{r}
# Prepare your data to make it suitable for a bar chart

vuln_2002_and_2007_top_6 <- vuln %>%
  
  filter(year %in% c(2002,2007)) %>%
  
  mutate(year = as.factor(year)) %>%
  
  select(country, year, vulnerable_pop) %>%
  
  group_by(year) %>%
  
  slice_max(order_by = vulnerable_pop, n = 6) %>%

  ungroup()

# To Display the data

vuln_2002_and_2007_top_6
```

```{r}
vuln_cluster <- ggplot(vuln_2002_and_2007_top_6, aes(x = country, y = vulnerable_pop, fill = year)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(
    title = "Top 6 countries by Vulnerable Population (2002 and 2007)",
    x = "Country",
    y = "Vulnerable Population"
  )

vuln_cluster
```

## Stacked bar chart

```{r}
# Prepare your data to make it suitable for a bar chart

vuln_1997_2002_2007_top_6 <- vuln %>%
  
  filter(year %in% c(1997,2002,2007)) %>%
  
  mutate(year = as.factor(year)) %>%
  
  select(country, year, vulnerable_pop) %>%
  
  group_by(year) %>%
  
  slice_max(order_by = vulnerable_pop, n = 6) %>%

  ungroup()

# To Display the data

vuln_1997_2002_2007_top_6
```

```{r}
vuln_stacked_bar <- ggplot(vuln_1997_2002_2007_top_6, aes(x = country, y = vulnerable_pop, fill = year)) +
  geom_bar(stat = "identity")

vuln_stacked_bar
```