---
title: "Basic Data Visualisation, How to make pie charts in R"
output: html_document
---

This block contains instructions on how to fill in this template.

**Please delete this block when creating your own file.**

This template should be used to create code snippet notebooks for each chapter in a learning resource. Follow the instructions below to complete the template: - Replace content in square brackets (e.g., [unit_name], [link], [chapter name]) with the relevant information. For example the title block below could be changed from "[unit_name], [chapter_name] - Code snippets" to "Basic Data Visualisation, Creating pie charts - Code snippets". - In the setup section add any instructions for users to follow before running the code snippets, such as installing packages or downloading data files. Add any necessary import statements or setup code in the first code cell. - Code snippets should be grouped into sections according to the sections in the learning resource. - Each code snippet from the learning resource should be placed in its own code cell under the relevant section, with any necessary context provided in a markdown block above each code cell. - Follow [tidyverse](https://style.tidyverse.org/) style guide for R code formatting. - Whenever you need to reference a file in [learning_resources](http://_vscodecontentref_/1), use:

``` r
file_path <- file.path(learning_resources_dir, "data", "filename.csv")
```

-   Definitions:
    -   "Unit name" refers to the unit name, which is also the name of the folder this python folder is in.
    -   "Chapter name" refers to the specific chapter that the code snippets are in. Chapters are the highest level headings in the learning resource.
    -   "Section name" refers to subheadings within chapters that code snippets are in. ---

## Basic Data Visualisation, How to make pie charts in R - Code examples

The following code relates to the **How to make pie charts in R** section of the **Basic Data Visualisation** learning resource on the [Analysis for Action](%5BPLACEHOLDER%5D) platform.

Use this code to get to grips with the basics of making pie charts in R as per the best practice covered in the written guidance. You are welcome to copy and adapt the code to meet your needs, but make sure that any outputs also align with any style requirements for your organisation.

The code applies each change separately and does not create any outputs.

## Troubleshooting

If you encounter any issues running the code snippets, consider the following troubleshooting steps: - Ensure you have followed the setup instructions in the [README.md](../../../README.md) file including `install.packages()` for required packages. - Run code from top to bottom to ensure all dependencies are met. - Restart your R session and clear all outputs. - For more detailed troubleshooting guidance, refer to the [troubleshooting document](%5BPLACEHOLDER%5D).

## Setup

Before running the code snippets, ensure you have: - followed the setup instructions in the [README.md](../../../README.md) file, - **[add further instructions as needed]**, - run the code block below.

```{r setup, message=FALSE}
# This chunk should include:
# * library() calls for required packages
# * paths to data files
# * any other setup required
library(readr)
library(here)
library(dplyr)
library(tidyr)
library(janitor)
library(ggplot2)


# Use here::here() to construct paths, do not setwd()
here::here()
vuln <- read.csv(here::here("learning_resources", "data", "vulnerable.csv"))
```

## Prepare your data

This code uses the 'vulnerable' dataset.

The raw data is too large to use in a pie chart. You need to filter it to make a smaller dataframe first.

```{r}
# Prepare your data to make it suitable for a bar chart

vuln_2007_top_6_pie <- vuln %>%
  # Filter the data to just the year 2007
  filter(year %in% ("2007")) %>%
  # Select just the country and vulnerable_pop columns
  select(country, vulnerable_pop) %>%
  # Select only the rows with the 6 highest vulnerable_pop value
  slice_max(order_by = vulnerable_pop, n = 6, with_ties = FALSE) %>%
  arrange(desc(vulnerable_pop)) %>%
  # Set factor levels so the pie stacks in the desired order
  mutate(country = factor(country, levels = country))


# To Display the data

vuln_2007_top_6_pie
```

## Make the base pie chart

```{r}
vuln_pie <- ggplot(vuln_2007_top_6_pie, aes(x = "", y = vulnerable_pop, fill = country)) +
  geom_col() +
  # start sets the location of the start of the first segment, 0 = vertical
  # direction sets the rotation on the pie chart, -1 = clockwise
  coord_polar("y", start = 0, direction = -1)

vuln_pie
```

# Remove breaks around the edge

```{r}
vuln_pie_clean <- ggplot(vuln_2007_top_6_pie, aes(x = "", y = vulnerable_pop, fill = country)) +
  geom_col() +
  coord_polar("y", start = 0, direction = -1) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank()
)

vuln_pie_clean
```

# Add white border between sectors

```{r}
vuln_pie_white_borders <- ggplot(vuln_2007_top_6_pie, aes(x = "", y = vulnerable_pop, fill = country)) +
  geom_col(colour = "#FFFFFF") +
  coord_polar("y", start = 0, direction = -1) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank()
)

vuln_pie_white_borders
```

# Change sector colours to ensure high contrast

```{r}
vuln_pie_colours <- ggplot(vuln_2007_top_6_pie, aes(x = "", y = vulnerable_pop, fill = country)) +
  geom_col(colour = "#FFFFFF") +
  coord_polar("y", start = 0, direction = -1) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank()
) +
  scale_fill_manual(
    values = c("#1E2A16", "#D17E38", "#444443", "#3EA27F", "#752111", "#618DD3")
)

vuln_pie_colours
```

# Add percentages as labels

```{r}
# Adapt data to include percentages
vuln_2007_top_6_pie_percent <- vuln_2007_top_6_pie %>%
  mutate(percent = vulnerable_pop / sum(vulnerable_pop))

vuln_2007_top_6_pie_percent
```

```{r}
vuln_pie_percent<- ggplot(vuln_2007_top_6_pie_percent, aes(x = 1, y = vulnerable_pop, fill = country)) +
  geom_col(colour = "#FFFFFF") +
  coord_polar("y", start = 0, direction = -1) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank()
) +
  scale_fill_manual(
    values = c("#1E2A16", "#D17E38", "#444443", "#3EA27F", "#752111", "#618DD3")
) +
  geom_text(
    aes(
      x = 1.3,
      label = percent(percent, accuracy = 0.01)),
    position = position_stack(vjust = 0.5),
    color = "#FFFFFF",
    size = 3,
  )

vuln_pie_percent
```

# Exporting

```{r}
vuln_pie_percent_big_text<- ggplot(vuln_2007_top_6_pie_percent, aes(x = 1, y = vulnerable_pop, fill = country)) +
  geom_col(colour = "#FFFFFF") +
  coord_polar("y", start = 0, direction = -1) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank()
) +
  scale_fill_manual(
    values = c("#1E2A16", "#D17E38", "#444443", "#3EA27F", "#752111", "#618DD3")
) +
  geom_text(
    aes(
      x = 1.3,
      label = percent(percent, accuracy = 0.01)),
    position = position_stack(vjust = 0.5),
    color = "#FFFFFF",
    # For the saved version the text size can be larger as it is a larger plot
    size = 6,
  )

ggsave(
  # Set the file name
  filename = "vuln_pie.svg",
  # Select the plot to save
  plot = vuln_pie_percent_big_text,
  # Set the save path
  path = "D:/Temp",
  # Set aspect ratio of plot, an aspect ratio of 16:9 will fill 1920x1080 displays
  width = 16,
  height = 9
)
```

